Modeling Interventions

### 2021-03-11 paper/paper.Rmd
VM - another pass

In physics, the "standard model" is a widely-accepted theory from the 1970s that describes all known elementary particles along with three of the four known fundamental forces in the universe. 
Although it is known to be incomplete and even incorrect in places, physics’ Standard Model nonetheless explains a wide variety of empirical phenomena, allowing scientists and engineers to model physical interactions with great precision and accuracy and to design tests that inform theory revision. 
Psychology in general has been criticized for lacking such formal theories that inform and drive empirical research [@muthukrishna2019problem]. 
We feel this critique can be applied fairly to the field of language acquisition.

@@

This basic framework corresponds nicely with a class of computational models that has been explored in the word learning literature: accumulator models. 
Our aim is to make this connection explicit, using item-response theory (IRT) from psychometric testing as a framework for connecting empirical data and previous accumulator models. 
Sitting at a level of analysis between generic regressions and cognitive process models, data-analytic cognitive models of this type offer a way to implement our verbal verbal theories and evaluate them quantitativelys. 
Importantly, the goal is not to produce 'the correct model," but rather to explore the ways that model assumptions lead to predictions about specific patterns of data. 
In fact, the greatest value of such models is the ability to identify the areas of greatest mismatch between data and model, highlighting areas of the theory requiring further investigation [@tauber2016].

@@

We are interested in children’s vocabulary, which we define as the set of words they know.^[
What makes a word is its own separate and difficult question, especially as we look across languages. Are "dog" and "dogs" two separate words? 
If not, what about "mouse" and "mice"? These questions are relatively inconsequential in morphologically simple languages like English and Mandarin; yet they are quite a bit more difficult in morphologically more complex languages like Turkish or even agglutinating languages like Inuktitut. 
We sidestep this issue by noting that the Turkish or Inuktitut learner is accumulating *something* – whether or not we call it a word *per se*, our hypothesis is that this unit is being accumulated and that its accumulation will be subject to many of the same dynamics and information sources as words are. 
For some initial data bearing out this supposition by comparing Turkish with other languages, see @braginsky2019.]
Driven by influential observations of the strong relationship between language input and children’s word learning [e.g., @huttenlocher1991early; @hart1995], computational modelers have created a range of hypotheses about how linguistic experiences drive learning. 
One class of models is *accumulator* models, which assume that linguistic experiences with words accumulate in separate registers (depicted as "buckets" in Figure 2), with those registers exceeding a particular threshold being learned.

@@

One difficulty in connecting such models to data is that the relevant variables are often measured in relative, rather than absolute, units. 
This practice is common, for example, measures of intelligence are given on a standardized scale defined by population variability. 
Yet in early language, we have access to absolute units of description, relatively unusually in a psychological domain. 
We can count how many words a child hears and express this as a rate [e.g., words per hour; @cristia2019], and we can similarly estimate how many words they know [e.g., by parents report of vocabulary size, @wordbankbook]. 
These absolute units give the potential for a model to make powerfully general quantitative predictions, which could be tested across different situations and populations [@dailey2021language]. 
As we describe below, however, few datasets have measurements of the relevant variables. 
This may be a case where theory outstrips data and further data collection efforts are necessary.

@@

Reasoning from first principles, you cannot learn the word "table if you do not hear it: input *must* predict learning. 
Yet correlational studies do not assess this relationship fully, and instead assess whether *variation* in input relates to *variation* in learning. 
Thus, a second line of work that comes closer to assessing the importance of language input focuses, not on differences between children, but on differences between words. These studies use regression models to predict which *words* are easier or more difficult, averaging across children [e.g., @goodman2008does; @braginsky2019]. 
Such studies show a strong association between word frequency and the age at which children typically acquire particular words, especially for object labels. 
This general finding provides convergent support for accumulator models, absent the confound of differences between individuals.

@@

This model outlines an approach similar to Item-Response Theory [IRT; @embretson2013item], an influential and popular psychometric modeling framework for understanding the performance of test-takers (in our case language learners) as they are assessed with a particular set of test items (particular words). 
In particular, IRT models jointly infer both how hard specific test items are and how accurate test takers are, recovering these latent parameters from the observed data. 
In the basic Rasch (or 1-parameter logistic) IRT model, a person i responds correctly to item j with probability determined by their ability ($\theta_i$) and the difficulty of item j ($d_j$):

@@

These parameters can easily be mapped onto the kind of accumulator model we have been describing: items are words (e.g. $d_j$ is the difficulty of word j), and $\theta_i$ is child $i$’s estimated latent language ability. 
In typical IRT models both item difficulties and person abilities are standardized (assumed to be normally-distributed and centered on 0) and unit-free. 
However, in principle, it is possible to map these scores to real-world distributions incorporating measured word frequencies and rates of children’s experienced input. 
With this mapping into absolute units, this kind of model can provide a quantitative linking hypothesis between measurements of input and learning. 
Recognizing the parallels between the standard model desiderata and IRT theory, we can use standard extensions of the 1PL (Rasch) IRT model to include both person-level and item-level covariates. 
Item-level (word-level) covariates of central interest are estimates of word frequency in child-directed speech, as well as lexical class [e.g., noun, predicate, or function word; @goodman2008does; @braginsky2019]. 
Person-level covariates of interest might include sex, socioeconomic status variables, or other demographic information.

This model as described above formalizes an important and testable claim: is the amount of language experience the important predictor of learning for a child, or are there other age-related factors beyond experience? 
Put simply, is a 2-year-old different than a 1-year-old, aside from having twice as much experience with each word? 
This theoretical question can be answered by simple model comparison between two versions of the standard model, one in which age is included as a person-level covariate, and one in which age is included as a scalar on the prevalence of each item (i.e., in which the frequency of "dog" for a 2-year-old is twice that of "dog" for a 1-year-old). 
Using the IRT framework to formalize theoretical questions and assess them via model comparison is a strength of this approach.

Below we report the results of this model comparison, fit to parent-reported vocabulary data from 5,492 English-speaking children in Wordbank. 
Each model estimated parameters for each child’s ability, and for the 680 words on the CDI Words and Sentences form. 
Covariates were the lexical class of each word (function word, adjective, noun, verb, or other), and the estimated number of monthly tokens per word a child hears (range: 0.2 - 19,286 tokens), based on estimates from the CHILDES corpus of child-directed speech [@macwhinney2000childes]. 
This model expects an average child to receive 1,200 words per hour, 12 hours each day, for a total of 438,300 tokens per month, of which 285,200 tokens (65.1%) are accounted for by tokens of CDI words.

In Model 1, we included age as a person-level covariate, allowing for interactions of age (in months), lexical class, and monthly tokens per word. 
Model 2 dropped the age covariate and instead scaled the expected tokens per month for each word by child age, yielding an estimate of lifetime tokens per word, and also allowed for interactions with lexical class. 
Model selection metrics (AIC and BIC) preferred the more complex Model 1, with a person-level age covariate separate from the item prevalence covariate. 
Predicted acquisition curves from Model 1 are shown in Figure 3, broken down by age, lexical class, and prevalence (left), revealing that nouns are learned more rapidly than verbs or function words, and that there are interactions of prevalence and word type (e.g., more frequent nouns are learned earlier, but there is little effect of prevalence for verbs and function words). 
Moreover, the model predicts item-level acquisition curves (Figure 3, right): e.g., "ball" is learned earlier than "dog", and "go" is easier than "have." 
Model 1 represents a new baseline standard model, which might be extended to include person-level groups such as SES and sex, as well as item-level covariates such as word length.

@@

The goal of a computational theory like the "standard model" described here is to describe the quantitative predictions that arise from a set of assumptions. 
Often, however, it is the *failure* of such a model to predict observed patterns of data that is most useful, as these failures point the way forward towards future refinements [@vanrooij2020theory]. 
We discuss three potential failures here.

**Leveraged learning and the role of processing**. As the model results above show, older children accumulate language from their input faster than younger children. 
There are at least three non-mutually exclusive potential reasons for this.
First, older children may be better at remembering their experiences due to independent developmental/maturational factors. 
Another possibility is that children *leverage* their knowledge of language to learn faster through any of several proposed mechanisms [@mitchell2009]. 
For example, if children reason about new words by exclusion [@markman1988children], they get better at learning new words just by having more potential items to exclude from consideration. 
A third possibility is that children are getting better at using the words they know. 
Several recent studies found that a surprising proportion of variance in the rate of young children’s vocabulary growth is accounted for by the speed with which individual children are able to process familiar words [e.g., @fernald2006picking; @marchman2016early].
Disentangling these distinct explanations will require careful measurement of both vocabulary and the potential mechanisms of leverage [cf. @lewis2020role].

**Beyond vocabulary**. The "standard model" proposed here models the accumulation of words.
Yet, of course, language is a rich, complex system in which words are inflected morphologically and composed syntactically to express compositional meanings. 
In early conceptions of language learning, syntax and morphology were thought of as distinct and unconnected, but this conception has not been borne out empirically. 
Instead, evidence from large-scale data shows again and again that the language system is "tightly woven," with extremely tight correlations between the acquisition of words, morphology, and syntax [@bornstein2013language,@wordbankbook]. 
Theoretically, accumulator models like the standard model are generic models of skill acquisition. 
Given the perspective that language learning is a form of skill acquisition [@chater2018language], such connections could lead the way towards extensions of the standard model to the accumulation of broader units of language like constructions.

**A theory for understanding acquisition in diverse contexts**. The "standard model" formalizes an implict assumption of the literature: namely, that early language learning follows the same general model for children in radically different circumstances. 
Yet this assumption could very well be false. 
Some children might learn more from overheard speech and others might learn more from child-directed speech [@sperry2019]. 
The kind of data necessary to test this assumption directly are only now being collected, for example, in studies that rigorously track learning outcomes and amount of language input in diverse populations [e.g., comparisons between children growing up in low-income, rural, indigenous communities and those in higher-income Western contexts; @casillas2020].

@@

An implicit theory drives much research and policy-making on early language acquisition: early language accumulates through a set of discrete experiences with individual words. 
The more experiences, the faster the words are learned. 
This implicit theory can be expressed as a simple computational model that makes quantitative predictions. 
By situating this model in a common psychometric framework, we show how it can be used to connect to large-scale, cross-linguistic datasets. 
This "standard model" synthesizes measures of language input and vocabulary growth, allowing us to formalize, test, and iteratively improve our theoretical understanding of early language acquisition. 
Moreover, this formalization allows us to identify specific gaps in current empirical approaches that must be closed to inform future theory development.

>>>>

In physics, the "standard model" is a widely-accepted theory from the 1970s that describes all known elementary particles along with three of the four known fundamental forces in the universe. 
Although it is known to be incomplete and even incorrect in places, physics’ Standard Model nonetheless explains a wide variety of empirical phenomena, allowing scientists and engineers to model physical interactions with great precision and accuracy and to design tests that inform theory revision. 
Psychology, in general, has been criticized for lacking such formal theories that inform and drive empirical research [@muthukrishna2019problem]. 
We feel this critique can be applied fairly to the field of language acquisition.

@@

This basic framework corresponds nicely with a class of computational models that has been explored in the word learning literature: accumulator models. 
Our aim is to make this connection explicit, using item-response theory (IRT) from psychometric testing as a framework for connecting empirical data and previous accumulator models. 
Sitting at a level of analysis between generic regressions and cognitive process models, data-analytic cognitive models of this type offer a way to implement our verbal theories and evaluate them quantitatively. 
Importantly, the goal is not to produce the "correct" model, but rather to explore the ways that model assumptions lead to predictions about specific patterns of data. 
In fact, the greatest value of such models is the ability to identify the areas of greatest mismatch between data and model, highlighting areas of the theory requiring further investigation [@tauber2016].

@@

We are interested in children’s vocabulary, which we define as the set of words they know.^[
What makes a word is its own separate and difficult question, especially as we look across languages. Are "dog" and "dogs" two separate words? 
If not, what about "mouse" and "mice"? These questions are relatively inconsequential in morphologically simple languages like English and Mandarin; yet they are quite a bit more difficult in morphologically more complex languages like Turkish or even agglutinating languages like Inuktitut. 
We sidestep this issue by noting that the Turkish or Inuktitut learner is accumulating *something* – whether or not we call it a word *per se*, our hypothesis is that this unit is being accumulated and that its accumulation will be subject to many of the same dynamics and information sources as words. 
For some initial data bearing out this supposition by comparing Turkish with other languages, see @braginsky2019.]
Driven by influential observations of the strong relationship between language input and children’s word learning [e.g., @huttenlocher1991early; @hart1995], computational modelers have created a range of hypotheses about how linguistic experiences drive learning. 
One class of models is *accumulator* models, which assume that linguistic experiences with words accumulate in separate registers (depicted as "buckets" in Figure 2), and that those registers exceeding a particular threshold are learned.

@@

One difficulty in connecting such models to data is that the relevant variables are often measured in relative, rather than absolute, units. 
This practice is common. For example, measures of intelligence are given on a standardized scale defined by population variability. 
Yet in early language, we have access to absolute units of description, relatively unusually in a psychological domain. 
We can count how many words a child hears and express this as a rate [e.g., words per hour; @cristia2019], and we can similarly estimate how many words they know [e.g., by parents' reports of vocabulary size, @wordbankbook]. 
These absolute units give the potential for a model to make powerfully general quantitative predictions, which could be tested across different situations and populations [@dailey2021language]. 
As we describe below, however, few datasets have measurements of the relevant variables. 
This may be a case where theory outstrips data and further data collection efforts are necessary.

@@

Reasoning from first principles, you cannot learn the word "table" if you do not hear it: input *must* predict learning. 
Yet correlational studies do not assess this relation fully, and instead assess whether *variation* in input relates to *variation* in learning. 
Thus, a second line of work that comes closer to assessing the importance of language input focuses, not on differences between children, but on differences between words. These studies use regression models to predict which *words* are easier or more difficult, averaging across children [e.g., @goodman2008does; @braginsky2019]. 
Such studies show a strong association between word frequency and the age at which children typically acquire particular words, especially for object labels. 
This general finding provides convergent support for accumulator models, absent the confound of differences between individuals.

@@

This model outlines an approach similar to Item-Response Theory [IRT; @embretson2013item], an influential and popular psychometric modeling framework for understanding the performance of test takers (in our case language learners) as they are assessed with a particular set of test items (particular words). 
In particular, IRT models jointly infer both the difficulty of specific test items and the accuracy of specific test takers, recovering these latent parameters from the observed data. 
In the basic Rasch (or 1-parameter logistic) IRT model, a person i responds correctly to item j with probability determined by their ability ($\theta_i$) and the difficulty of item j ($d_j$):

@@

These parameters can easily be mapped onto the accumulator model we have been describing: items are words (e.g. $d_j$ is the difficulty of word j), and $\theta_i$ is child $i$’s estimated latent language ability. 
In typical IRT models, both item difficulties and person abilities are standardized (assumed to be normally-distributed and centered on 0) and unit-free. 
However, in principle, it is possible to map these scores to real-world distributions incorporating measured word frequencies and rates of children’s experienced input. 
With this mapping into absolute units, this kind of model can provide a quantitative linking hypothesis between measurements of input and learning. 
Recognizing the parallels between the standard model desiderata and IRT theory, we can use standard extensions of the 1PL (Rasch) model to include both item-level and person-level covariates. 
Item-level (word-level) covariates of central interest are estimates of word frequency in child-directed speech, as well as lexical class [e.g., noun, predicate, or function word; @goodman2008does; @braginsky2019]. 
Person-level covariates of interest might include sex, socioeconomic status variables, or other demographic information.

This model as described above formalizes an important and testable claim: is the amount of language experience the important predictor of learning for a child, or are there other age-related factors beyond experience? 
Put simply, is a 2-year-old different than a 1-year-old, aside from having twice as much experience with each word? 
This theoretical question can be answered by simple model comparison between two versions of the standard model, one in which age is a person-level covariate, and one in which age is a scalar on the prevalence of each item (i.e., the frequency of "dog" for a 2-year-old is twice that of a 1-year-old). 
Using the IRT framework to formalize theoretical questions and assess them via model comparison is a strength of this approach.

We report the results of this model comparison, fit to parent-reported vocabulary data from 5,492 English-speaking children in Wordbank. 
Each model estimated parameters for each child’s ability, and for the 680 words on the CDI Words and Sentences form [@fenson2007]. 
Covariates were each word's lexical class (function word, adjective, noun, verb, or other) and the estimated number of monthly tokens a child hears (range: 0.2 - 19,286 tokens), based on child-directed speech from CHILDES [@macwhinney2000childes]. 
This model expects an average child to receive 1,200 words/hour, 12 hours/day, for a total of 438,300 tokens/month, of which 285,200 tokens (65.1%) are  tokens of CDI words.

In Model 1, we included age as a person-level covariate, allowing for interactions of age (in months), lexical class, and monthly tokens per word. 
Model 2 dropped the age covariate and instead scaled the expected tokens/month for each word by child age, yielding an estimate of lifetime tokens per word and also allowed for interactions with lexical class. 
Model selection metrics (AIC and BIC) preferred the more complex Model 1. 
Predicted acquisition curves from Model 1 are shown in Figure 3 by age, lexical class, and prevalence (left), revealing that nouns are learned more rapidly than verbs or function words, and that there are interactions of prevalence and word type (e.g., more frequent nouns are learned earlier, but there is little effect of prevalence for verbs and function words). 
Moreover, the model predicts item-level acquisition curves (Figure 3, right): e.g., "ball" is learned earlier than "dog" and "go" is easier than "have." 
Model 1 represents a new baseline standard model, which might be extended to include person-level groups such as SES and sex, as well as item-level covariates such as word length.

@@

The goal of a computational theory like the "standard model" is to describe the quantitative predictions that arise from a set of assumptions. 
Often, however, it is the *failure* of such a model to predict observed patterns of data that is most useful, as these failures point the way forward towards future refinements [@vanrooij2020theory]. 
We discuss three potential failures here.

**Leveraged learning and the role of processing**. As the model results show, older children accumulate language from their input faster than younger children. 
There are at least three non-mutually exclusive potential reasons for this.
First, older children may be better at remembering their experiences due to independent developmental/maturational factors. 
Another possibility is that children *leverage* their knowledge of language to learn faster through any of several proposed mechanisms [@mitchell2009]. 
For example, if children reason about new words by exclusion [@markman1988children], they get better at learning new words just by having more potential items to exclude from consideration. 
A third possibility is that children are getting better at using the words they know. 
Several recent studies found that a surprising proportion of variance in the rate of young children’s vocabulary growth is accounted for by the speed with which individual children are able to process familiar words [e.g., @fernald2006picking; @marchman2016early].
Disentangling these distinct explanations will require careful measurement of both vocabulary and the potential mechanisms of leverage [cf. @lewis2020role].

**Beyond vocabulary**. The "standard model" proposed here models the accumulation of words.
Yet, of course, language is a rich, complex system in which words are inflected morphologically and composed syntactically to express compositional meanings. 
In early views, syntax and morphology were conceptualized as distinct and unconnected, but this conception has not been borne out empirically. 
Instead, evidence shows again and again that the language system is "tightly woven," with extremely tight correlations between the acquisition of words, morphology, and syntax [@bornstein2013language,@wordbankbook]. 
Theoretically, accumulator models like the standard model are generic models of skill acquisition. 
If language learning is a form of skill acquisition [@chater2018language], such connections could lead the way towards extensions of the standard model to the accumulation of broader units of language-like constructions.

**A theory for understanding acquisition in diverse contexts**. The "standard model" formalizes an implict assumption: namely, that early language learning follows the same general model for all children, even those in radically different circumstances. 
Yet this assumption could very well be false. 
Some children might learn more from overheard speech and others might learn more from child-directed speech [@sperry2019]. 
The kind of data necessary to test this assumption directly are only now being collected, for example, in studies that rigorously track learning outcomes and amount of language input in diverse populations [e.g., comparisons between children in low-income, rural, indigenous communities and those in higher-income Western contexts; @casillas2020].

@@

An implicit theory drives much research and policy-making on early language acquisition: early language accumulates through discrete experiences with individual words. 
The more experiences, the faster the words are learned. 
This implicit theory can be expressed as a simple computational model that makes quantitative predictions. 
By situating this model in a common psychometric framework, we show how it can be used to connect to large-scale, cross-linguistic datasets. 
This "standard model" synthesizes measures of language input and vocabulary growth, allowing us to formalize, test, and iteratively improve our theoretical understanding of early language acquisition. 
Moreover, this formalization allows us to identify specific gaps in current empirical approaches that must be closed to inform future theory development.
